---
title: "Modeling Soil Organic Carbon in Southeast Asia: Methods and Climate Implication"
author: "Tram M. Tran"
date: "2025–2026"
institute: "Denison University"
format:
  html:
    toc: true
    toc-depth: 3
    theme: lux
    code-tools: true
    code-fold: true
    code-summary: "Show code"
    code-copy: hover
    link-external-newwindow: true
    tbl-cap-location: top
    fig-cap-location: bottom
self-contained: true
editor: source
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center",
  out.width  = "90%",
  results    = "hold",
  fig.show   = "hold",
  error      = TRUE,
  warning    = FALSE,
  message    = FALSE
)
par(mar = c(4.1, 4.1, 1.1, 4.1))

# Collapsible output helper
hooks <- knitr::knit_hooks$get()
hook_foldable <- function(type) {
  force(type)
  function(x, options) {
    res <- hooks[[type]](x, options)
    if (isFALSE(options[[paste0("fold.", type)]])) return(res)
    paste0(
      "<details open><summary>",
      gsub("^p", "P", gsub("^o", "O", type)),
      "</summary>\n\n", res, "\n\n</details>"
    )
  }
}
knitr::knit_hooks$set(
  output = hook_foldable("output"),
  plot   = hook_foldable("plot")
)
```

# Import Libraries

All packages used in this analysis are loaded here. `set.seed(123)` ensures
reproducibility of the Random Forest models.

```{r libraries}
# Data wrangling
library(dplyr)
library(tidyr)
library(stringr) #for extraction of string in depth_class
library(readr)
library(sf) #coordinates mapping
library(terra) # extract from raster file
library(ggmap) # mapping      
library(ggplot2) #visual
library(patchwork)   # for panel figures
library(randomForest) #RF
library(lme4)        #Linear Mixed Effects Model
library(lmerTest)    # p-values for lmer fixed effects
library(MuMIn)       # pulling R-square
library(emmeans)     # estimated marginal means
library(piecewiseSEM)

set.seed(123) # Reproducibility of Random Forest
```

---

# 1. Data Import & Cleaning

## 1.1 Load raw SOC dataset

The dataset is from Gomez et al. (2024), 4,341 observations across nine Southeast Asian countries (1987–2023). See [doi: 10.1038/s41597-024-03213-3](https://doi.org/10.1038/s41597-024-03213-3).

```{r load-data}
soc_raw <- read.csv("Data/clean1.csv", stringsAsFactors = FALSE)

# Making sure coordinates are numeric
soc_raw$lon <- as.numeric(soc_raw$Longtitude)
soc_raw$lat <- as.numeric(soc_raw$Latitude)

#Checking rows and columns
cat("Raw dataset dimensions:", nrow(soc_raw), "rows x", ncol(soc_raw), "cols\n") #print data
```

## 1.2 Merge CHELSA climate data

Mean annual temperature (bio1, Celsius) and mean annual precipitation (bio12, mm) are extracted from CHELSA v2.1 climatologies (1981–2010) using site coordinates. See [Karger et al. (2021)](https://doi.org/10.16904/envidat.228).

```{r chelsa}
# Convert to simple features for spatial extraction. Tutorial: https://cran.r-project.org/web/packages/sf/vignettes/sf1.html 
pts_sf <- st_as_sf(soc_raw, coords = c("lon", "lat"), crs = 4326)

# Load CHELSA rasters (pre-downloaded to ../CHELSA_*)
r_bio1  <- rast("../CHELSA_bio1_1981-2010_V.2.1.tif") #Raster file
r_bio12 <- rast("../CHELSA_bio12_1981-2010_V.2.1.tif") #Raster file

bio <- c(r_bio1, r_bio12)
names(bio) <- c("bio1_temp", "bio12_precip")

# Extract values and merge to data "soc_climate"
# tutorial: https://rdrr.io/cran/terra/man/extract.html 
vals <- terra::extract(bio, vect(pts_sf))
soc_climate <- cbind(soc_raw, vals)

#Checking rows after merging CHELSA
cat("After CHELSA merge:", nrow(soc_climate), "rows\n") 
```

## 1.3 Depth classification

Depth strings are parsed into a midpoint value and classified into three strata:
0–30 cm (topsoil), 30–60 cm (midsoil), and >60 cm (deepsoil), following the
rationale that ANOVA confirms a significant SOC × depth interaction (see Section 3.1).

```{r depth-class}
# understanding depth str_extract:
# ^ = start of string, [0-9] one or more digits. $ = end of string --> grab the first and last number of string
soc_climate <- soc_climate %>%
  mutate(
    depth_low  = as.numeric(str_extract(depth, "^[0-9]+")),
    depth_high = as.numeric(str_extract(depth, "[0-9]+$")),
    depth_mid  = (depth_low + depth_high) / 2, # taking the average as the main depth to categorize
    depth_class = case_when(
      depth_mid <= 30  ~ "0-30",
      depth_mid > 30 & depth_mid <= 60 ~ "30-60",
      depth_mid > 60  ~ "60+",
      TRUE  ~ NA_character_ ))
table(soc_climate$depth_class,useNA = "ifany") #there are 22 NAs
```

## 1.4 SOC unit harmonisation & log-transformation

SOC values reported in diverse units are converted to g/kg (concentration) and Mg/ha (stock). This analysis uses **g/kg** (≈ 62% of observations). Due to strong right-skew and outliers, log-transformation is applied to all SOC values prior to modelling.

```{r soc-units}
soc_climate <- soc_climate %>%
  mutate(
    # Concentration (g/kg)
    SOC_gKg = case_when(
      soc.value_units == "%"     ~ soc.value * 10,
      soc.value_units == "g_kg"  ~ soc.value,
      soc.value_units == "g_g"   ~ soc.value * 1000,
      soc.value_units == "mg_g"  ~ soc.value,        # 1 mg/g = 1 g/kg
      soc.value_units == "mg_kg" ~ soc.value / 1000,
      TRUE ~ NA_real_
    ),
    # Stock (Mg/ha)
    SOC_MgHa = case_when(
      soc.value_units == "Mg_ha" ~ soc.value,
      soc.value_units == "g_m2"  ~ soc.value * 0.01,
      soc.value_units == "kg_m2" ~ soc.value * 10,
      soc.value_units == "kg_ha" ~ soc.value / 1000,
      TRUE ~ NA_real_
    )
  )

# Log-transform (shift if needed to ensure all values > 0)
safe_log <- function(x) {
  minv <- min(x[is.finite(x)], na.rm = TRUE)
  shift <- ifelse(minv <= 0, abs(minv) + 1e-6, 0)
  log(x + shift)
}

soc_climate <- soc_climate %>%
  mutate(
    log_SOC_gKg  = safe_log(SOC_gKg),
    log_SOC_MgHa = safe_log(SOC_MgHa)
  )

print(summary(soc_climate$SOC_gKg))
print(summary(soc_climate$log_SOC_gKg))
```

## 1.5 Clean analysis-ready dataset

Apply inclusion criteria: retain observations with valid g/kg SOC, depth class, management descriptor, and climate data. Scale temperature and precipitation for model comparability.

```{r soc-main}
soc_main <- soc_climate %>%
  mutate(
    depth_class          = factor(depth_class, levels = c("0-30", "30-60", "60+")),
    main_descriptor      = factor(main_descriptor),
    secondary_descriptor = factor(secondary_descriptor),
    additional_descriptor = factor(additional_descriptor),
    year                 = as.numeric(year),
    country              = as.factor(country),
    temp_s               = as.numeric(scale(bio1_temp)),
    precip_s             = as.numeric(scale(bio12_precip))
  ) %>%
  filter(
    !is.na(log_SOC_gKg),
    !is.na(depth_class),
    !is.na(secondary_descriptor),
    !is.na(temp_s),
    !is.na(precip_s)
  )

# 2699 observations we work with
print(table(soc_main$depth_class))
```

## 1.6 Create management intensity variables

Binary intensity variables are constructed for fertilization (1 = with, 0 = without) and tillage (1 = with, 0 = without), encoding treatment presence based on `additional_descriptor` variable.

```{r intensity-vars}
soc_main <- soc_main %>%
  mutate(
    # Fertilization: 1 = fertilized, 0 = unfertilized
    fert_intensity = case_when(
      secondary_descriptor == "Fertilization" &
        str_detect(additional_descriptor, regex("without", ignore_case = TRUE)) ~ 0,
      secondary_descriptor == "Fertilization" &
        str_detect(additional_descriptor, regex("with",    ignore_case = TRUE)) ~ 1,
      secondary_descriptor == "Fertilization" ~ 1,  # default: treated
      TRUE ~ NA_real_ ),
    # Tillage: 1 = tilled (any intensity), 0 = no tillage
    tillage_intensity = case_when(
      secondary_descriptor == "Tillage" &
        str_detect(additional_descriptor, regex("No tillage",          ignore_case = TRUE)) ~ 0,
      secondary_descriptor == "Tillage" &
        str_detect(additional_descriptor, regex("Minimum|Reduced|Conventional", ignore_case = TRUE)) ~ 1,
      secondary_descriptor == "Tillage" ~ 1,         # default: tilled
      TRUE ~ NA_real_))
```

## 1.7 Create subsets by depth

```{r depth-subsets}
depth_030  <- filter(soc_main, depth_class == "0-30")
depth_3060 <- filter(soc_main, depth_class == "30-60")
depth_60   <- filter(soc_main, depth_class == "60+")
```

---

# 2. Descriptive Statistics & Exploratory Analysis

## 2.1 Study site map

```{r map-sites, fig.cap="Figure 1. Study locations across Southeast Asia. Point colour indicates depth class; size reflects number of co-located observations. Countries with the most observations are Indonesia, Thailand, and Vietnam."}

map_data <- soc_main %>%
  filter(!is.na(lon), !is.na(lat)) %>%
  count(lon, lat, country, name = "n_obs")

bbox <- c(left = 92, bottom = -11, right = 141, top = 28)

map_attempt <- 
  ggplot() +
    borders("world",
            regions = c("Indonesia","Thailand","Vietnam","Philippines",
                        "Malaysia","Cambodia","Laos","Myanmar",
                        "Singapore","Brunei","Timor-Leste"),
            fill = "grey85", colour = "white") +
    geom_point(data = map_data,
               aes(lon, lat, size = n_obs),
               color = "#c0392b", alpha = 0.65) +
    coord_fixed(xlim = c(92, 141), ylim = c(-11, 28)) +
    labs(title = "Study locations for SOC measurements in Southeast Asia",
         x = "Longitude", y = "Latitude") +
    scale_size(range = c(1, 6), name = "Number of Observations per Location") +
    theme_minimal()

map_attempt
```

## 2.2 SOC summary statistics by depth class

```{r soc-summary}
soc_main %>%
  group_by(depth_class) %>%
  summarise(
    n         = n(),
    mean_gkg  = round(mean(SOC_gKg,     na.rm = TRUE), 2),
    median_gkg = round(median(SOC_gKg,  na.rm = TRUE), 2),
    sd_gkg    = round(sd(SOC_gKg,       na.rm = TRUE), 2),
    min_gkg   = round(min(SOC_gKg,      na.rm = TRUE), 2),
    max_gkg   = round(max(SOC_gKg,      na.rm = TRUE), 2),
    mean_log  = round(mean(log_SOC_gKg, na.rm = TRUE), 3),
    median_log = round(median(log_SOC_gKg, na.rm = TRUE), 3)
  )
```

## 2.3 ANOVA: depth class effect on SOC

A one-way ANOVA tests whether mean log-SOC differs significantly across the
three depth strata. This justifies running all subsequent models separately
per depth class.

```{r anova-depth, fig.cap="Figure 2. Distribution of log-transformed SOC (g/kg) by depth class. Topsoil consistently shows higher SOC; deepsoil shows the widest spread."}
# Filter to complete g/kg observations
gkg <- soc_main %>% filter(!is.na(log_SOC_gKg))

# One-way ANOVA
aov_depth <- aov(log_SOC_gKg ~ depth_class, data = gkg)
summary(aov_depth)
```


```{r anova-depth, fig.cap="Figure 2. Distribution of log-transformed SOC (g/kg) by depth class. Topsoil consistently shows higher SOC; deepsoil shows the widest spread."}
# Boxplot
ggplot(gkg, aes(x = depth_class, y = log_SOC_gKg, fill = depth_class)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.85) +
  scale_fill_manual(values = c("0-30" = "#7570b3",
                               "30-60" = "#d95f02",
                               "60+"   = "#1b9e77")) +
  labs(title = "Distribution of SOC by Depth Class",
       x = "Depth (cm)", y = "log(SOC) g/kg") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5),
        panel.grid.minor = element_blank())
```

**Result:** F(`r df.residual(aov_depth)` residual df) = `r round(summary(aov_depth)[[1]][["F value"]][1], 1)`, p < 0.001 — SOC concentrations differ significantly across depth strata, justifying depth-stratified modelling.

## 2.4 Climate variable distributions

```{r climate-dist, fig.cap="Figure 3. Distributions of mean annual precipitation (mm) and temperature (°C) across all observations. Most sites receive 1,000–3,000 mm precipitation and experience 24–28 °C."}
soc_main %>%
  select(bio1_temp, bio12_precip) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(variable = recode(variable,
    bio1_temp    = "Mean Annual Temperature (°C)",
    bio12_precip = "Mean Annual Precipitation (mm)")) %>%
  ggplot(aes(x = value, fill = variable)) +
  geom_histogram(bins = 30, colour = "white", alpha = 0.85) +
  facet_wrap(~variable, scales = "free") +
  scale_fill_manual(values = c(
    "Mean Annual Temperature (°C)"    = "#E64B35",
    "Mean Annual Precipitation (mm)"  = "#4DBBD5")) +
  labs(title = "Distribution of Climate Variables", x = NULL, y = "Frequency") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 13),
        plot.title = element_text(face = "bold", hjust = 0.5),
        panel.grid.minor = element_blank())
```

## 2.5 SOC distribution by land use and depth

```{r landuse-dist, fig.cap="Figure 4. Log-transformed SOC (g/kg) by land-use type (Annual, Agroforestry, Perennial) and depth class. Perennial systems tend to maintain higher SOC at midsoil and deepsoil depths."}
soc_landuse <- soc_main %>%
  filter(main_descriptor == "Land use", !is.na(log_SOC_gKg)) %>%
  select(id, year, lon, lat, country, main_descriptor,
         secondary_descriptor, temp_s, precip_s, depth_class, log_SOC_gKg)

ggplot(na.omit(soc_landuse),
       aes(x = secondary_descriptor, y = log_SOC_gKg,
           fill = secondary_descriptor)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.85) +
  facet_wrap(~depth_class) +
  scale_fill_manual(values = c(
    "Agroforestry" = "#8B4513",
    "Annual"       = "#2E7D32",
    "Perennial"    = "#DAA520")) +
  labs(title = "Distribution of SOC by Land Use and Depth",
       x = "Land Use", y = "log(SOC) g/kg") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 13),
        plot.title = element_text(face = "bold", hjust = 0.5),
        panel.grid.minor = element_blank())
```

---

# 3. Random Forest: Predictive Importance

Random Forest models are fit separately per depth stratum using temperature, precipitation, and management category (`secondary_descriptor`) as predictors of log(SOC). Variable importance is assessed via permutation-based %IncMSE.

## 3.1 Fit models (ntree = 500)

```{r rf-models}
# Helper: strip NAs needed for RF
rf_data <- function(df) {
  df %>% filter(
    !is.na(log_SOC_gKg),
    !is.na(temp_s), !is.na(precip_s),
    !is.na(secondary_descriptor)
  )
}

rf_d1 <- randomForest(
  log_SOC_gKg ~ temp_s + precip_s + secondary_descriptor,
  data = rf_data(depth_030), importance = TRUE, ntree = 500)

rf_d2 <- randomForest(
  log_SOC_gKg ~ temp_s + precip_s + secondary_descriptor,
  data = rf_data(depth_3060), importance = TRUE, ntree = 500)

rf_d3 <- randomForest(
  log_SOC_gKg ~ temp_s + precip_s + secondary_descriptor,
  data = rf_data(depth_60), importance = TRUE, ntree = 500)
```

## 3.2 Model performance (R² and RMSE)

```{r rf-performance}
rf_metrics <- function(model, label) {
  r2   <- 1 - model$mse[model$ntree] / var(model$y, na.rm = TRUE)
  rmse <- sqrt(model$mse[model$ntree])
  data.frame(Depth = label, R2 = round(r2, 3), RMSE = round(rmse, 3))
}

rf_perf <- bind_rows(
  rf_metrics(rf_d1, "0-30 cm"),
  rf_metrics(rf_d2, "30-60 cm"),
  rf_metrics(rf_d3, "60+ cm")
)
print(rf_perf)
```

## 3.3 Predicted vs. observed — panel figure

```{r rf-pred-obs, fig.cap="Figure 5. Random Forest predicted vs. observed log(SOC g/kg) by depth class. Red line = 1:1 reference. Topsoil (R² = 0.88) shows the tightest fit; midsoil (R² = 0.74) shows the most dispersion."}
pred_df <- bind_rows(
  data.frame(Observed = rf_data(depth_030)$log_SOC_gKg,
             Predicted = predict(rf_d1), Depth = "0–30 cm"),
  data.frame(Observed = rf_data(depth_3060)$log_SOC_gKg,
             Predicted = predict(rf_d2), Depth = "30–60 cm"),
  data.frame(Observed = rf_data(depth_60)$log_SOC_gKg,
             Predicted = predict(rf_d3), Depth = "60+ cm")
)
pred_df$Depth <- factor(pred_df$Depth, levels = c("0–30 cm","30–60 cm","60+ cm"))

ggplot(pred_df, aes(x = Observed, y = Predicted)) +
  geom_point(colour = "#4DBBD5", alpha = 0.4, size = 0.9) +
  geom_abline(slope = 1, intercept = 0, colour = "red", linewidth = 0.8) +
  facet_wrap(~Depth, scales = "free") +
  labs(title = "Random Forest Model Fit — Predicted vs. Observed log(SOC)",
       x = "Observed log(SOC g/kg)", y = "Predicted log(SOC g/kg)") +
  theme_minimal(base_size = 13) +
  theme(strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold", hjust = 0.5))
```

## 3.4 Variable importance — panel figure

```{r rf-varimp, fig.cap="Figure 6. Variable importance (%IncMSE) by depth class. Precipitation is the leading predictor at topsoil; temperature dominates at midsoil and deepsoil. Management importance declines with depth."}
imp_label <- function(model, depth_col) {
  data.frame(
    Variable   = rownames(randomForest::importance(model)),
    Importance = randomForest::importance(model)[, 1],
    Depth      = depth_col
  ) %>%
    mutate(Variable = recode(Variable,
      temp_s               = "Temperature",
      precip_s             = "Precipitation",
      secondary_descriptor = "Management"))
}

imp_all <- bind_rows(
  imp_label(rf_d1, "0–30 cm"),
  imp_label(rf_d2, "30–60 cm"),
  imp_label(rf_d3, "60+ cm")
)
imp_all$Depth <- factor(imp_all$Depth, levels = c("0–30 cm","30–60 cm","60+ cm"))

depth_colours <- c("0–30 cm" = "#7570b3", "30–60 cm" = "#d95f02", "60+ cm" = "#1b9e77")

ggplot(imp_all, aes(x = reorder(Variable, Importance), y = Importance,
                     fill = Depth)) +
  geom_col(width = 0.5) +
  coord_flip() +
  facet_wrap(~Depth) +
  scale_fill_manual(values = depth_colours) +
  labs(title = "Variable Importance in SOC Prediction (Random Forest)",
       x = NULL, y = "% Increase in MSE") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold", hjust = 0.5))
```

---

# 4. Linear Mixed-Effects Models

LMMs quantify the direction and magnitude of fixed-effect predictors on
log(SOC) while controlling for country and year as crossed random intercepts.
Agroforestry is the reference category for land-use comparisons.

**Model structure:**

$$
\log(\text{SOC}_{i}) =
\beta_0 + \beta_1 \text{LandManagement}_{i}
+ \beta_2 \text{Temp}_{i} + \beta_3 \text{Precip}_{i}
+ u_{\text{country}[i]} + v_{\text{year}[i]} + \varepsilon_i
$$

where $u_{\text{country}} \sim \mathcal{N}(0, \sigma^2_{\text{country}})$,
$v_{\text{year}} \sim \mathcal{N}(0, \sigma^2_{\text{year}})$,
$\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$.

## 4.1 Land-use models by depth

```{r lmm-landuse}
# Agroforestry as reference
relevel_lu <- function(df) {
  df %>%
    filter(main_descriptor == "Land use") %>%
    mutate(secondary_descriptor = relevel(
      factor(secondary_descriptor), ref = "Agroforestry"))
}

m_lu_030  <- lmer(log_SOC_gKg ~ secondary_descriptor + temp_s + precip_s +
                    (1|country) + (1|year),
                  data = relevel_lu(depth_030), REML = FALSE, na.action = na.omit)

m_lu_3060 <- lmer(log_SOC_gKg ~ secondary_descriptor + temp_s + precip_s +
                    (1|country) + (1|year),
                  data = relevel_lu(depth_3060), REML = FALSE, na.action = na.omit)

m_lu_60   <- lmer(log_SOC_gKg ~ secondary_descriptor + temp_s + precip_s +
                    (1|country) + (1|year),
                  data = relevel_lu(depth_60), REML = FALSE, na.action = na.omit)

cat("--- Topsoil (0–30 cm): Land Use ---\n");  print(summary(m_lu_030))
cat("--- Midsoil (30–60 cm): Land Use ---\n"); print(summary(m_lu_3060))
cat("--- Deepsoil (60+ cm): Land Use ---\n");  print(summary(m_lu_60))

# Marginal and conditional R²
cat("\nR² (marginal | conditional):\n")
cat("0-30 cm: "); print(round(r.squaredGLMM(m_lu_030), 3))
cat("30-60cm: "); print(round(r.squaredGLMM(m_lu_3060), 3))
cat("60+ cm:  "); print(round(r.squaredGLMM(m_lu_60), 3))

# 95% Wald CIs
cat("\n95% Wald CIs — 0-30 cm:\n");  print(confint(m_lu_030,  method = "Wald"))
cat("95% Wald CIs — 30-60 cm:\n"); print(confint(m_lu_3060, method = "Wald"))
cat("95% Wald CIs — 60+ cm:\n");   print(confint(m_lu_60,   method = "Wald"))
```

## 4.2 Fertilization models by depth

```{r lmm-fert}
m_fert_030  <- lmer(log_SOC_gKg ~ fert_intensity + temp_s + precip_s +
                      (1|country) + (1|year),
                    data = depth_030, REML = FALSE, na.action = na.omit)

m_fert_3060 <- lmer(log_SOC_gKg ~ fert_intensity + temp_s + precip_s +
                      (1|country) + (1|year),
                    data = depth_3060, REML = FALSE, na.action = na.omit)

m_fert_60   <- lmer(log_SOC_gKg ~ fert_intensity + temp_s + precip_s +
                      (1|country) + (1|year),
                    data = depth_60, REML = FALSE, na.action = na.omit)

cat("--- Topsoil (0–30 cm): Fertilization ---\n");  print(summary(m_fert_030))
cat("--- Midsoil (30–60 cm): Fertilization ---\n"); print(summary(m_fert_3060))
cat("--- Deepsoil (60+ cm): Fertilization ---\n");  print(summary(m_fert_60))

cat("\n95% Wald CIs — 0-30 cm:\n");  print(confint(m_fert_030,  method = "Wald"))
cat("95% Wald CIs — 30-60 cm:\n"); print(confint(m_fert_3060, method = "Wald"))
cat("95% Wald CIs — 60+ cm:\n");   print(confint(m_fert_60,   method = "Wald"))
```

## 4.3 Tillage models by depth

Tillage at 60+ cm is not modelled due to n = 2 observations with tillage treatment.

```{r lmm-tillage}
m_till_030  <- lmer(log_SOC_gKg ~ tillage_intensity + temp_s + precip_s +
                      (1|country) + (1|year),
                    data = depth_030, REML = FALSE, na.action = na.omit)

m_till_3060 <- lmer(log_SOC_gKg ~ tillage_intensity + temp_s + precip_s +
                      (1|country) + (1|year),
                    data = depth_3060, REML = FALSE, na.action = na.omit)

cat("--- Topsoil (0–30 cm): Tillage ---\n");  print(summary(m_till_030))
cat("--- Midsoil (30–60 cm): Tillage ---\n"); print(summary(m_till_3060))

cat("\n95% Wald CIs — 0-30 cm:\n");  print(confint(m_till_030,  method = "Wald"))
cat("95% Wald CIs — 30-60 cm:\n"); print(confint(m_till_3060, method = "Wald"))
```

## LMM fixed effects

The forest plot below uses the manually curated coefficient table (from model
outputs above) for clean visual presentation, consistent with the manuscript.

```{r forest-plot, fig.cap="Figure 7. LMM fixed-effect estimates (β) with 95% confidence intervals by management predictor and depth class. Filled points = significant (CI does not cross zero). Dashed line at zero."}
lmm_coefs <- tribble(
  ~Name,                 ~Slope,  ~LCI,   ~UCI,   ~Significant,
  "0-30 Annual",         -0.44,  -0.67,  -0.21,  "Yes",
  "30-60 Annual",        -1.29,  -1.85,  -0.74,  "Yes",
  "60+ Annual",          -0.22,  -0.53,   0.08,  "No",
  "0-30 Perennial",      -0.48,  -0.69,  -0.27,  "Yes",
  "30-60 Perennial",     -0.83,  -1.40,  -0.26,  "Yes",
  "60+ Perennial",       -0.23,  -0.52,   0.07,  "No",
  "0-30 Tillage",        -0.02,  -0.16,   0.11,  "No",
  "30-60 Tillage",       -0.15,  -0.42,   0.12,  "No",
  "0-30 Fertilization",   0.19,   0.09,   0.29,  "Yes",
  "30-60 Fertilization",  0.30,  -0.14,   0.61,  "No",
  "60+ Fertilization",    0.40,  -0.003,  0.80,  "No"
) %>%
  separate(Name, into = c("Depth", "Practice"), sep = " ", remove = FALSE) %>%
  mutate(
    Depth       = factor(Depth, levels = c("60+", "30-60", "0-30")),
    Significant = factor(Significant, levels = c("Yes", "No")),
    Name        = factor(Name, levels = rev(c(
      "0-30 Annual",       "0-30 Perennial",    "0-30 Tillage",    "0-30 Fertilization",
      "30-60 Annual",      "30-60 Perennial",   "30-60 Tillage",   "30-60 Fertilization",
      "60+ Annual",        "60+ Perennial",                        "60+ Fertilization"
    )))
  )

ggplot(lmm_coefs, aes(y = Name, x = Slope, colour = Depth)) +
  geom_errorbarh(aes(xmin = LCI, xmax = UCI), height = 0.25, linewidth = 0.8) +
  geom_point(aes(size = Significant)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("0-30" = "#7570b3",
                                 "30-60" = "#d95f02",
                                 "60+"   = "#1b9e77")) +
  scale_size_manual(values = c("Yes" = 4, "No" = 2)) +
  labs(title = "Effect of Land Management on SOC by Depth Class",
       subtitle = "Points = estimate | Horizontal lines = 95% CI",
       x = "Slope Estimate (log g/kg)", y = NULL,
       colour = "Depth (cm)", size = "Significant?") +
  theme_minimal(base_size = 14) +
  theme(axis.text.y  = element_text(size = 11),
        plot.title   = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

---

# 5. Piecewise Structural Equation Modeling (SEM)

piecewiseSEM fits locally-estimated LMM sub-models to simultaneously test direct (Climate → SOC) and indirect (Climate → Management → SOC) pathways. Country is included as a random effect in each sub-model. SEM models are depth-stratified for fertilization and tillage; land-use effects are assessed via Estimated Marginal Means (Section 6).

## 5.1 Fertilization pathways

### Topsoil (0–30 cm)

```{r sem-fert-030}
sem_fert_030 <- psem(
  lmer(fert_intensity ~ temp_s + precip_s + (1|country), data = depth_030),
  lmer(log_SOC_gKg   ~ fert_intensity + temp_s + precip_s + (1|country),
       data = depth_030)
)
summary(sem_fert_030)
plot(sem_fert_030)
```

### Midsoil (30–60 cm)

```{r sem-fert-3060}
sem_fert_3060 <- psem(
  lmer(fert_intensity ~ temp_s + precip_s + (1|country), data = depth_3060),
  lmer(log_SOC_gKg   ~ fert_intensity + temp_s + precip_s + (1|country),
       data = depth_3060)
)
summary(sem_fert_3060)
plot(sem_fert_3060)
```

### Deepsoil (60+ cm)

```{r sem-fert-60}
sem_fert_60 <- psem(
  lmer(fert_intensity ~ temp_s + precip_s + (1|country), data = depth_60),
  lmer(log_SOC_gKg   ~ fert_intensity + temp_s + precip_s + (1|country),
       data = depth_60)
)
summary(sem_fert_60)
plot(sem_fert_60)
```

## 5.2 Tillage pathways

Tillage at 60+ cm is excluded (n = 2 with tillage treatment, insufficient for
piecewiseSEM stability).

### Topsoil (0–30 cm)

```{r sem-till-030}
sem_till_030 <- psem(
  lmer(tillage_intensity ~ temp_s + precip_s + (1|country), data = depth_030),
  lmer(log_SOC_gKg       ~ tillage_intensity + temp_s + precip_s + (1|country),
       data = depth_030)
)
summary(sem_till_030)
```

### Midsoil (30–60 cm)

```{r sem-till-3060}
sem_till_3060 <- psem(
  lmer(tillage_intensity ~ temp_s + precip_s + (1|country), data = depth_3060),
  lmer(log_SOC_gKg       ~ tillage_intensity + temp_s + precip_s + (1|country),
       data = depth_3060)
)
summary(sem_till_3060)
```

---

# 6. Land-Use Effects: Estimated Marginal Means

Because land-use categories are nominal and cannot serve as mediating response variables in piecewiseSEM, their climate-adjusted effects on SOC are estimated via Estimated Marginal Means (EMMs) from the LMMs fitted in Section 4.1, holding temperature and precipitation at their means.

## 6.1 Compute EMMs for each depth

```{r emm-compute}
emm_030  <- emmeans(m_lu_030,  ~ secondary_descriptor)
emm_3060 <- emmeans(m_lu_3060, ~ secondary_descriptor)
emm_60   <- emmeans(m_lu_60,   ~ secondary_descriptor)

cat("EMMs — Topsoil (0–30 cm):\n");  print(summary(emm_030))
cat("\nEMMs — Midsoil (30–60 cm):\n"); print(summary(emm_3060))
cat("\nEMMs — Deepsoil (60+ cm):\n");  print(summary(emm_60))
```

## 6.2 Panel figure: EMMs across all depths

```{r emm-panel, fig.cap="Figure 9. Estimated Marginal Means (EMMs) for land-use categories at each depth class, holding temperature and precipitation at their means. Points = emmean; bars = 95% CI. Note convergence of EMMs at topsoil (p = 0.8) and divergence at midsoil."}
lu_colours <- c(
  "Agroforestry" = "#8B4513",
  "Annual"       = "#2E7D32",
  "Perennial"    = "#DAA520"
)

make_emm_plot <- function(emm_obj, depth_label) {
  df <- as.data.frame(emm_obj)
  ggplot(df, aes(x = secondary_descriptor, y = emmean,
                 colour = secondary_descriptor)) +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2,
                  linewidth = 0.8) +
    geom_text(aes(label = round(emmean, 2)),
              hjust = 1.5, vjust = -0.5, size = 3.5) +
    scale_colour_manual(values = lu_colours) +
    labs(title = depth_label,
         x = "Land Use", y = "Predicted log(SOC g/kg)") +
    theme_minimal(base_size = 13) +
    theme(legend.position  = "none",
          plot.title       = element_text(face = "bold", hjust = 0.5),
          panel.grid.minor = element_blank())
}

p1 <- make_emm_plot(emm_030,  "0–30 cm (p = 0.8, ns)")
p2 <- make_emm_plot(emm_3060, "30–60 cm")
p3 <- make_emm_plot(emm_60,   "60+ cm")

# patchwork panel
(p1 | p2 | p3) +
  plot_annotation(
    title   = "SOC (g/kg) by Land Use Type and Depth",
    subtitle = "EMMs from LMM, climate covariates held at their mean values",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14, hjust = 0.5),
      plot.subtitle = element_text(colour = "grey40", hjust = 0.5)
    )
  )
```

---

# 7. Session Info

```{r session-info}
sessionInfo()
```
